Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-01 paper.tex
Various changes implementing Sally's comments

\begin{abstract}
Recombination is an important and ongoing feature of SARS-CoV-2 evolution,
and presents major challenges to [things important to public health].
% ... genomic surveillance programs, predictive models for symptoms and severity, policy recommendations such as vaccine effectiveness and booster frequency, etc. (https://doi.org/10.1016/S1473-3099(22)00831-3)
Detection of new recombinant strains is challenging, with methods
[being a bit rubbish].
% ... with few methods that facilitate rapid and systematic (reproducible) classification .https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/
In addition, many basic approaches fundamental to the study of viral
evolution assume that recombination is negligible, in that a single
phylogenetic tree can represent the genetic ancestry of the
circulating strains. Here we present an initial version of
\texttt{sc2ts}, a method to automatically detect recombinants
in real time and to cohesively integrate them into a
genealogy in the form of an ancestral recombination graph (ARG),
which jointly records mutation, recombination and genetic
inheritance. We infer two ARGs under
different sampling strategies, and study their properties.
One contains 1.27 million sequences
sampled up to June 30, 2021, and the second is more sparsely sampled,
consisting of 784K sequences sampled up to June 30, 2022.
We find that both ARGs are
% better words?
highly consistent with
known features of SARS-CoV-2 evolution, recovering the basic
backbone phylogeny, mutational spectra, and recapitulating
details on the majority of known recombinant lineages.
Using the well-established and feature-rich \texttt{tskit} library,
the ARGs can also be stored concisely and processed efficiently
using standard Python tools. For example, the ARG for 1.27 million
sequences---encoding the inferred reticulate ancestry,
genetic variation, and extensive metadata---requires
58MB of storage,
and loads in less than a second.
%JK Not sure if we'd actually say it like this, but this is the
% basic gist I'd like to get accross[
The ability to fully integrate the effects of recombination into
downstream analyses, to quickly and automatically detect new recombinants,
and to provide an efficient and convenient platform for computation
makes \texttt{sc2ts} a promising approach, worthy of
further investment and development.
\end{abstract}

\section{Introduction}
% Recombination is an important force in SARS-CoV-2, recombinants
% have arisen and they have spread to high frequencies
Recombination via template switching is a common feature
of the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genetic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of more
immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
% Early in the SARS-CoV-2 pandemic there was insufficient
% genetic variation for meaningful recombination.
Early on in the COVID-19 pandemic, the levels of genetic diversity
were too low to precipitate the emergence of distinctive recombinant strains.
By late 2020, however, the appearance and
spread of variants of concern (VoC), designated into classes such as Alpha and
Delta which harboured multiple characteristic mutations,
created the conditions required for recombination to occur,
and recombinant strains with onward transmission became detectable at
low frequencies~\citep{Jackson2021-ik}.
More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for
recombination~\citep{Bal2022-hq}.
In early 2023, multiple recombinant lineages have successfully
established and spread to high frequency, and accounting for recombinant
ancestry is now essential in understanding the evolution of SARS-CoV-2.

@@

% [PARA 4: we present sc2ts based on these recent advances. Summarise
% results, with forward refs to Results sections.]
Here we present a preliminary version of \texttt{sc2ts}, a novel method
for inferring ARGs for SARS-CoV-2 at pandemic scale, in real time.
Building on the open-source \texttt{tskit} library, the method
explicitly reconstructs genealogies with both
mutation and recombination, which
can be conveniently and efficiently analysed using standard Python data science
tools. Inference is based on incrementally adding batches of sequences
based on their collection dates, and proceeds in three phases.
% First, likely attachment nodes in the current ARG
% (or paths, if inferred to be a recombinant) are
First, possible paths connecting each sample to the current ARG
are inferred (allowing for recombination) using the
% chosen for each of the samples in the daily batch using the
Li and Stephens (LS) model (Section~\ref{sec:ls}). Then, since many samples
in the batch can share an attachment path,
we infer phylogenetic trees for each of the clusters separately using standard
methods (Section~\ref{sec:sample-cluster-tree-inference}).
Finally, we attach the trees for these sample-clusters to the current
ARG, and apply some parsimony-based heuristics to address issues
introduced by the inherent greediness of this strategy
(Section~\ref{sec:parsimony-heuristics}).
Using the current preliminary version of \texttt{sc2ts}, we infer two
large ARGs (with 1,265,685 and 657,239 samples, respectively), and study
the properties of these ARGs to illustrate the power of the method
and to inform subsequent development. We find that these
ARGs accurately reflect known patterns of phylogenetic relationships
between samples (Section~\ref{sec:backbone_phylogeny}),
mutational spectra (Section~\ref{sec:mutation_spectrum}),
and automatically identify the majority of known recombinant
lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),
with an unprecedented degree of precision in the
genomic location of recombination breakpoints
(Section~\ref{sec:breakpoint_intervals})
and the relationship between parental sequences
(Section~\ref{sec:parent_divergence}).
We hope that these benefits of accurate joint estimation of
genetic inheritance with mutation and recombination events will generate community
interest and development of the \texttt{sc2ts} method, and
more generally in applying the efficient and mature software
of the \texttt{tskit} ecosystem to pandemic-scale SARS-CoV-2 data.

@@

\subsection{Divergence between recombinant parents}
\label{sec:parent_divergence}
% Paragraph 1, what does all this mean?
% Purpose of this paragraph - remind the reader that we can only do this
% sort of things because we have an ARG, and we can do  this at a far
% greater level of granularity than existing methods.
An important advantage of explicitly inferring an ARG compatible
with a given sample of sequences (Section~\ref{sec:args})
is that this allows us to precisely trace the origins of recombinants.
Methods based on motifs among pre-specified lineages
provide us with estimates of the origins of recombinants as
parents belonging to one or other of these groups. Defining the
groups requires care [FIXME and other fiddly things to mention Katherine?]
% ... Defining the groups is a highly subjective process, involving numerous manual steps that require subject matter expertise of the organism, that often use web sites as data sources which are difficult to version-control, and thus hinder reproducibility. The web paper by the PH4GE group is one of the best sources for the reality (and pain) of how it's done: https://pha4ge.org/resource/identifying-sars-cov-2-recombinants/
RIPPLES~\citep{Turakhia2022-it} does provide precise estimates
of the parental origins of a recombinant, but since it depends
on an existing phylogenetic tree, cannot reason about
nested recombinant origins.

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
\caption{\label{fig:recomb_mrcas}
Date of common ancestry between the parents on either side of recombination
breakpoints, as a function of the divergence time between the parents.
MRCAs of parents associated with Pango designated recombinants (XA, etc)
are identified in orange. Larger symbols represent breakpoints in recombination
nodes ancestral to five or more samples.
Horizontal dotted lines show the four most common MRCA nodes,
which tend to be associated with major outbreaks and with many immediate children.
The stacked histogram shows the distribution of parental divergence times, ranging
from parents that have diverged only a handful of days previously, to much more
divergent parent lineages. See Figure~\ref{fig:recomb_mrcas_voc_breakdown}
for equivalent plots broken down by parental VoC classification.
}
\end{figure}

% Paragraph 2, what do we plot, specifically?
In this section we explore the detailed ancestral relationships between
recombinant parents by investigating the patterns of divergence between
them.
We focus on the Long ARG because it covers time
periods where substantial recombination is known to have occurred,
and contains samples from 33 Pango X lineages (i.e., those inferred
to have recombinant ancestry; see Section~\ref{sec:pango_x_lineages} for
further analysis).
Figure~\ref{fig:recomb_mrcas} shows the estimated date of the most recent
common ancestor (MRCA) of the parent nodes for each recombination breakpoint,
plotted against the divergence between these parents (i.e., the
total branch length from the parents to their MRCA in the trees to
the immediate left and right of the breakpoint).
As in Section~\ref{sec:breakpoint_intervals}, in these plots we
exclude breakpoints associated with ``singleton'' recombination nodes
(those ancestral to only one sample).
Larger points distinguish those breakpoints which occur in nodes
ancestral to 5 or more samples, comprising 316 breakpoints from
291 recombination nodes. The criterion of 5 descendants matches the
minimum number required to designate a new Pango recombinant lineage
\citep{Rambaut2020-dw}.
Note that as each plotted point represents a breakpoint, a recombination
node with more than one breakpoint (e.g., with 3 or more parents,
comprising ${\sim}10 \%$ of the
recombination nodes in
Figure~\ref{fig:recomb_mrcas}) will be represented by several points.

% Paragraph 3, what do we say about the dates/nodes of the MRCAs?
The date of the MRCA of recombinant parents is
concentrated in several banded rows in Figure~\ref{fig:recomb_mrcas}. These are
largely due to a few MRCAs shared by many
recombinants (the top four are indicated in the figure).
These shared MRCAs lie near the root of large expansions
% Anyone reading this will know what these are
% (B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
% and BA.2: second Omicron wave)
% as demonstrated by the number of samples with those
% Pango designations traback to each ancestral node.
% The majority of these frequently-shared MRCAs
and the majority are associated with large polytomies, likely
indicating a rapid and under-sampled expansion of a clade.

% Paragraph 4, what do we say about the divergence times?
Figure~\ref{fig:recomb_mrcas} shows that there is a
large range in divergence times among recombinant parents, with a broad
spread of times from about 10 to 80 weeks prior to the recombination event
and a minor additional peak of recombinants that diverged
${\sim}95$ weeks, corresponding to recombinants whose
parental common ancestor traces to early 2020. This latter group should
contain, for example Delta-Omicron recombinants. More widely, we can
further classify the breakpoints by the VoC combinations of their
parent lineages. Considering only the Alpha, Delta, and Omicron VoC classes,
such a classification reveals that the majority of breakpoints have two Delta or
two Omicron parents, and that Omicron and Delta are the variants associated
with the most recombination (Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
This may reflect either sampling intensity, the
prevalence of cases (which increases the
chance of coinfection and recombination), or possible heterogeneity in
recombination probabilities among lineages.

The time estimates in these figures should be
treated with a degree of caution,
since non-sample nodes are crudely dated in the current version
of \texttt{sc2ts} (see Section~\ref{sec:node_dating}, in particular
for discussion of how these dates might be improved using existing
methods).
Nevertheless, it is clear
that \texttt{sc2ts} can identify recombination between lineages
that are only a few weeks diverged.

\subsection{Recombinant Pango lineages}
\label{sec:pango_x_lineages}
In this section we focus on the detailed ancestry of samples that have been
previously identified as recombinants, i.e., designated as belonging
to a Pango lineage with a name starting with an ``X''.
We focus primarily on the Long ARG
since it contains many more recombinants (the status
of Pango X lineages in the Wide ARG is briefly
summarised in Section~\ref{sec-pango_x_wide_arg}).
Designation of samples to Pango lineages is not a straightforward task,
and there can be significant variation between methods.
Here, we consider two different assignments of Pango lineages to samples in
the Long ARG, those provided by Nextclade and by GISAID.
% GISAID lineage calls come from the pangoLEARN algorithm.
Using the Nextclade assignments, the Long ARG contains 749 samples from
33 Pango X lineages (711 samples from 33 lineages when we
remove singleton recombinants).
In contrast, using the GISAID designations we
have 515 samples from 38 X lineages
(511 samples from 35 lineages when we filter singleton recombinants). Of the GISAID designations,
28 are shared with Nextclade (26 after filtering singletons).
This variation in classifications highlights the uncertainty that exists when
assigning Pango X lineages to samples, and is important to keep in mind
when interpreting the results here.

We focus here on the 749 Nextclade-designated recombinant samples, all
but two of which descend from a recombination node in the Long ARG.
These two samples are designated XP, and a likely
explanation for the absence of a corresponding recombination event in the Long ARG
is that the characteristic multibase deletion for XP
(\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
during preprocessing (See Section~\ref{sec:data_preprocessing} for details and potential
improvements). Of the remaining X designated samples, 38 are singleton recombinants,
descending
from recombination nodes that are ancestral only to that sample (10 are XZ; 6
are XE;  3 each from XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1
sample from each of XB, XM, XJ, XAF, XAH and XAJ). Such samples are likely to
be enriched for sequencing errors and lineage designation artefacts, and so we
exclude them from further analysis in this section.

@@

\caption{\label{fig:pango-simple-origin-graph} Examples of non-nested
Nextclade Pango X lineages. (A) Subgraph for XA in the Long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Vertical position of nodes does not
show absolute time, but simply relative rank (parents above children). Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by \texttt{sc2ts}, whose
Pango status is imputed. Genomic regions inherited by the recombination node are shown;
breakpoints correspond to the maximum breakpoint position inferred by \texttt{sc2ts}.
(B) Equivalent subgraph for the 17 XAG samples in the Long ARG, with abbreviated labelling. Where
non-XAG samples are ancestral to further unplotted samples, the number of unplotted descendant samples
is marked as ``+1'', ``+7'', etc.
(C) Equivalent subgraphs for both origination events involving the four XD samples in the Long ARG.
Details of the mutations and sample node identities for all three plots are provided in
supplementary Figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}, which also provide alternative GISAID Pango designations.
} \end{figure}

One of the simplest examples is the case of XA, corresponding to group A
of \citet{Jackson2021-ik} as discussed above. Figure~\ref{fig:pango-simple-origin-graph}A
shows the exact relationships inferred by \texttt{sc2ts} as a subgraph of the Long ARG.
Here, paths have been traced from all
Nextclade-identified XA samples (in red) to the closest other sample nodes in the ARG.
Sample nodes are plotted as larger circles, but the subgraph also includes intermediate,
non-sample nodes: i.e., inserted nodes such as UPGMA, recombination, mutation collapse, and
reversion push nodes, plotted as smaller circles. Dotted lines show where this subgraph links
to the rest of the ARG. Above recombination nodes, only ancestral nodes are shown, meaning
that the subgraph is not extended to show additional descendants of recombinant parents.

It is clear from the XA subgraph that all the samples labelled XA by Nextclade trace to a
single originating recombination node, whose genome is a composite of a B.1.177.18 lineage
on the left of the genome and a B.1.1.7 lineage on the right. In the subgraph we show
the maximum genomic position for the recombination breakpoint, here at position 22,227
(corresponding to a breakpoint strictly less than 22,228, see Table~\ref{tab:jackson})

@@

The classification of originating recombination events is dependent on
accurate designation of Pango lineages to samples. It is therefore important
to note that if the GISAID Pango designations are used, many of the samples
marked here as XAB are reclassified as BA.2 (and XAG becomes fully monophyletic,
see supplementary Figure \ref{fig:pango_XAG_gisaid_graph}). This is an
independent confirmation of the uncertainty in designation of these XAG-related
samples.

Six of the 16 Nextclade designated lineages (XA, XAC, XAE, XF, XK, and XS) are of the
basic (XA) type with no other Pango designations among their descendants.
The remaining ten are of the XAG type with multiple Pango
designated lineages as additional descendants of the originating recombination event.
In some cases,
these may, however, be a result of erroneous Pango designations.
Table~\ref{tab:pango-recombinants} also shows the official
Pango designated parent lineages and \texttt{sc2ts} inferred parent lineages,
which extensively agree (although \texttt{sc2ts} provides more a precise
parent designation).

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. Using Nextclade designations, 15 Pango X lineages are
inferred to have multiple
recombinant origins, such that their samples trace to more than one most recent
recombination node in the ARG. Of these, 11 are cases where the
recombinants are independent rather than nested (i.e., there is no overlap in the list
of descendant samples for each recombination node). Most have a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend, and which agrees with the offical Pango designated parent lineages
(see Table~\ref{tab:pango-recombinants}, but note that
in XJ and XU there are too few Pango X samples to decide on a ``main'' clade).

>>>>

\begin{abstract}
Recombination is an ongoing and increasingly important feature of circulating lineages of SARS-CoV-2, challenging how we represent the evolutionary history of this virus and giving rise to new variants of potential public health concern by combining transmission and immune evasion properties of different lineages. Detection of new recombinant strains is challenging, with most methods looking for breaks between sets of mutations that characterize distinct lineages. 
In addition, many basic approaches fundamental to the study of viral
evolution assume that recombination is negligible, in that a single
phylogenetic tree can represent the genetic ancestry of the
circulating strains. Here we present an initial version of
\texttt{sc2ts}, a method to automatically detect recombinants
in real time and to cohesively integrate them into a
genealogy in the form of an ancestral recombination graph (ARG),
which jointly records mutation, recombination and genetic
inheritance. We infer two ARGs under
different sampling strategies, and study their properties.
One contains 1.27 million sequences
sampled up to June 30, 2021, and the second is more sparsely sampled,
consisting of 784K sequences sampled up to June 30, 2022.
We find that both ARGs are
% better words?
highly consistent with
known features of SARS-CoV-2 evolution, recovering the basic
backbone phylogeny, mutational spectra, and recapitulating
details on the majority of known recombinant lineages.
Using the well-established and feature-rich \texttt{tskit} library,
the ARGs can also be stored concisely and processed efficiently
using standard Python tools. For example, the ARG for 1.27 million
sequences---encoding the inferred reticulate ancestry,
genetic variation, and extensive metadata---requires
58MB of storage,
and loads in less than a second.
%JK Not sure if we'd actually say it like this, but this is the
% basic gist I'd like to get accross[
The ability to fully integrate the effects of recombination into
downstream analyses, to quickly and automatically detect new recombinants,
and to provide an efficient and convenient platform for computation
makes \texttt{sc2ts} a promising approach, worthy of
further investment and development.
\end{abstract}

\section{Introduction}
% Recombination is an important force in SARS-CoV-2, recombinants
% have arisen and they have spread to high frequencies
Recombination via template switching is a common feature
of the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genetic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of more
immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
Early on in the COVID-19 pandemic, the levels of genetic diversity
were too low to precipitate the emergence of distinctive recombinant strains.
By late 2020, however, the appearance and
spread of variants of concern (VoC), designated into classes such as Alpha and
Delta which harboured multiple characteristic mutations,
created the conditions required for recombination to occur,
and recombinant strains with onward transmission became detectable at
low frequencies~\citep{Jackson2021-ik}.
More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for
recombination~\citep{Bal2022-hq}.
In early 2023, multiple recombinant lineages have successfully
established and spread to high frequency, and accounting for recombinant
ancestry is now essential in understanding the evolution of SARS-CoV-2.

@@

% [PARA 4: we present sc2ts based on these recent advances. Summarise
% results, with forward refs to Results sections.]
Here we present a preliminary version of \texttt{sc2ts}, a novel method
for inferring ARGs for SARS-CoV-2 at pandemic scale, in real time.
Building on the open-source \texttt{tskit} library, the method
explicitly reconstructs genealogies with both
mutation and recombination, which
can be conveniently and efficiently analysed using standard Python data science
tools. As illustrated in Figure~\ref{fig:overview_sc2ts}, inference is based on incrementally adding batches of sequences
based on their collection dates, and proceeds in three phases.
% First, likely attachment nodes in the current ARG
% (or paths, if inferred to be a recombinant) are
First, possible paths connecting each sample to the current ARG
are inferred (allowing for recombination) using the
% chosen for each of the samples in the daily batch using the
Li and Stephens (LS) model (Figure~\ref{fig:overview_sc2ts}A, B; Section~\ref{sec:ls}). Then, since many samples
in the batch can share an attachment path,
we infer phylogenetic trees for each of the clusters separately using standard
methods (Figure~\ref{fig:overview_sc2ts}C; Section~\ref{sec:sample-cluster-tree-inference}).
Finally, we attach the trees for these sample-clusters to the current
ARG, and apply some parsimony-based heuristics to address issues
introduced by the inherent greediness of this strategy
(Figure~\ref{fig:overview_sc2ts}D, E; Section~\ref{sec:parsimony-heuristics}).
Using the current preliminary version of \texttt{sc2ts}, we infer two
large ARGs (with 1,265,685 and 657,239 samples, respectively), and study
the properties of these ARGs to illustrate the power of the method
and to inform subsequent development. We find that these
ARGs accurately reflect known patterns of phylogenetic relationships
between samples (Section~\ref{sec:backbone_phylogeny}),
mutational spectra (Section~\ref{sec:mutation_spectrum}),
and automatically identify the majority of known recombinant
lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),
with an unprecedented degree of precision in the
genomic location of recombination breakpoints
(Section~\ref{sec:breakpoint_intervals})
and the relationship between parental sequences
(Section~\ref{sec:parent_divergence}).
We hope that these benefits of accurate joint estimation of
genetic inheritance with mutation and recombination events will generate community
interest and development of the \texttt{sc2ts} method, and
more generally in applying the efficient and mature software
of the \texttt{tskit} ecosystem to pandemic-scale SARS-CoV-2 data.

@@

\subsection{Divergence between recombinant parents}
\label{sec:parent_divergence}
% Paragraph 2, what do we plot, specifically?
In this section we explore the detailed ancestral relationships between
recombinant parents by investigating the patterns of divergence between
them.
We focus on the Long ARG because it covers time
periods where substantial recombination is known to have occurred
and contains samples from 33 Pango X lineages (i.e., those inferred
to have recombinant ancestry; see Section~\ref{sec:pango_x_lineages} for
further analysis).
Figure~\ref{fig:recomb_mrcas} shows the estimated date of the most recent
common ancestor (MRCA) of the parent nodes for each recombination breakpoint,
plotted against the divergence between these parents (i.e., the
total branch length from the parents to their MRCA in the trees to
the immediate left and right of the breakpoint).
As in Section~\ref{sec:breakpoint_intervals}, in these plots we
exclude breakpoints associated with ``singleton'' recombination nodes
(those ancestral to only one sample).
Larger points distinguish those breakpoints which occur in nodes
ancestral to 5 or more samples, comprising 316 breakpoints from
291 recombination nodes. The criterion of 5 descendants matches the
minimum number required to designate a new Pango recombinant lineage
\citep{Rambaut2020-dw}.
Note that as each plotted point represents a breakpoint, a recombination
node with more than one breakpoint (e.g., with 3 or more parents,
comprising ${\sim}10 \%$ of the
recombination nodes in
Figure~\ref{fig:recomb_mrcas}) will be represented by several points.

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/recombination_node_mrcas.pdf}
\caption{\label{fig:recomb_mrcas}
Date of common ancestry between the parents on either side of recombination
breakpoints, as a function of the divergence time between the parents.
MRCAs of parents associated with Pango designated recombinants (XA, etc)
are identified in orange. Larger symbols represent breakpoints in recombination
nodes ancestral to five or more samples.
Horizontal dotted lines show the four most common MRCA nodes,
which tend to be associated with major outbreaks and with many immediate children.
The stacked histogram shows the distribution of parental divergence times, ranging
from parents that have diverged only a handful of days ago, to much more
divergent parent lineages. See Figure~\ref{fig:recomb_mrcas_voc_breakdown}
for equivalent plots broken down by parental VoC classification.
}
\end{figure}

% Paragraph 3, what do we say about the dates/nodes of the MRCAs?
The date of the MRCA of recombinant parents is
concentrated in several banded rows in Figure~\ref{fig:recomb_mrcas}. These are
largely due to a few MRCAs shared by many
recombinants (the top four are indicated in the figure).
These shared MRCAs lie near the root of large expansions,
% Anyone reading this will know what these are
% (B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
% and BA.2: second Omicron wave)
% as demonstrated by the number of samples with those
% Pango designations traback to each ancestral node.
% The majority of these frequently-shared MRCAs
and the majority are associated with large polytomies, likely
indicating a rapid and under-sampled expansion of a clade (e.g. major
Delta and Omicron waves).

% Paragraph 4, what do we say about the divergence times?
Figure~\ref{fig:recomb_mrcas} shows that there is a
large range in divergence times among parents of detected recombinants, with a broad
spread of times from about 10 to 80 weeks prior to the recombination event
and a minor additional peak involving recombination between lineages that diverged
${\sim}95$ weeks ago, corresponding to common ancestors which trace back to early 2020.
This latter group should
contain, for example Delta-Omicron recombinants. More widely, we can
further classify the breakpoints by the VoC combinations of their
parent lineages. Considering only the Alpha, Delta, and Omicron VoC classes,
such a classification reveals that the majority of breakpoints have two Delta or
two Omicron parents, and that Omicron and Delta are the variants associated
with the most recombination (Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
This may reflect either sampling intensity, the
prevalence of cases (which increases the
chance of coinfection and recombination), or possible heterogeneity in
recombination probabilities among lineages.

The time estimates in these figures should be
treated with a degree of caution,
because non-sample nodes are crudely dated in the current version
of \texttt{sc2ts} (see Section~\ref{sec:node_dating}, in particular
for discussion of how these dates might be improved using existing
methods).
Nevertheless, it is clear
that \texttt{sc2ts} can identify recombination between lineages
that are only a few weeks diverged.

\subsection{Recombinant Pango lineages}
\label{sec:pango_x_lineages}
In this section we focus on the detailed ancestry of samples that have been
previously identified as recombinants, i.e., designated as belonging
to a Pango lineage with a name starting with an ``X''.
We focus primarily on the Long ARG
which contains many more recombinants (the status
of Pango X lineages in the Wide ARG is briefly
summarised in Section~\ref{sec-pango_x_wide_arg}).
Designation of samples to Pango lineages is not a straightforward task,
and there can be significant variation between methods.
Here, we consider two different assignments of Pango lineages to samples in
the Long ARG, those provided by Nextclade and by GISAID.
% GISAID lineage calls come from the pangoLEARN algorithm.
Using the Nextclade assignments, the Long ARG contains 749 samples from
33 Pango X lineages (711 samples from 33 lineages when we
remove singleton recombinants).
In contrast, using the GISAID designations we
have 515 samples from 38 X lineages
(511 samples from 35 lineages when we filter singleton recombinants). Of the GISAID designations,
28 are shared with Nextclade (26 after filtering singletons).
This variation in classifications highlights the uncertainty that exists when
assigning Pango X lineages to samples, and is important to keep in mind
when interpreting the results here.

We focus here on the 749 Nextclade-designated recombinant samples, all
but two of which descend from a recombination node in the Long ARG.
These two exceptions are both designated XP, and a likely
explanation for the absence of a corresponding recombination event in the Long ARG
is that the characteristic multibase deletion for XP
(\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
during our preprocessing (see Section~\ref{sec:data_preprocessing} for details and potential
improvements). Of the remaining X designated samples, 38 are singleton recombinants,
descending
from recombination nodes that are ancestral only to that sample (10 are labelled XZ; 6
are XE;  3 each from XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1
sample from each of XB, XM, XJ, XAF, XAH and XAJ). Such samples are likely to
be enriched for sequencing errors and lineage designation artefacts, and so we
exclude them from further analysis in this section.

@@

\caption{\label{fig:pango-simple-origin-graph} Examples of non-nested
Nextclade Pango X lineages. (A) Subgraph for XA in the Long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Vertical position of nodes does not
show absolute time, but simply relative rank (parents above children). Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by \texttt{sc2ts}, whose
Pango status is imputed. Genomic regions inherited by the recombination node are shown;
breakpoints correspond to the rightmost breakpoint position inferred by \texttt{sc2ts}.
(B) Equivalent subgraph for the 17 XAG samples in the Long ARG, with abbreviated labelling. Where
non-XAG samples are ancestral to further unplotted samples, the number of unplotted descendant samples
is marked as ``+1'', ``+7'', etc.
(C) Equivalent subgraphs for both origination events involving the four XD samples in the Long ARG.
Details of the mutations and sample node identities for all three plots are provided in
supplementary Figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}, which also provide alternative GISAID Pango designations.
} \end{figure}

One of the simplest examples is the case of XA, corresponding to group A
of \citet{Jackson2021-ik} as discussed above. Figure~\ref{fig:pango-simple-origin-graph}A
shows the exact relationships inferred by \texttt{sc2ts} as a subgraph of the Long ARG.
Here, paths have been traced from all
Nextclade-identified XA samples (in red) to the closest other sample nodes in the ARG.
Sample nodes are plotted as larger circles, but the subgraph also includes intermediate,
non-sample nodes (i.e. inserted by sc2ts, see Sections~\ref{sec:sample-cluster-tree-inference} and \ref{sec:parsimony-heuristics}). Dotted lines show where this subgraph links
to the rest of the ARG. Above recombination nodes, only ancestral nodes are shown, meaning
that the subgraph is not extended to show additional descendants of recombinant parents.

It is clear from the XA subgraph that all the samples labelled XA by Nextclade trace to a
single originating recombination node, whose genome is a composite of a B.1.177.18 lineage
on the left of the genome and a B.1.1.7 lineage on the right. In the subgraph we show
the rightmost genomic position for the recombination breakpoint, here at position 22,227
(corresponding to a breakpoint strictly less than 22,228, see Table~\ref{tab:jackson})

@@

The classification of originating recombination events is dependent on
accurate designation of Pango lineages to samples. It is therefore important
to note that if the GISAID Pango designations are used, many of the samples
marked here as XAB are reclassified as BA.2 and XAG becomes fully monophyletic
(although still not an immediate descendant of the originating recombination node,
see Supplementary Figure \ref{fig:pango_XAG_gisaid_graph}). This is an
independent confirmation of the uncertainty in designation of these XAG-related
samples.

Six of the 16 Nextclade designated lineages (XA, XAC, XAE, XF, XK, and XS) are of the
basic (XA) type with no other Pango designations among their descendants.
The remaining ten are of the XAG type with multiple Pango
designated lineages as additional descendants of the originating recombination event.
In some cases,
these may, however, be a result of erroneous Pango designations.
Table~\ref{tab:pango-recombinants} also shows the official
Pango designated parent lineages and \texttt{sc2ts} inferred parent lineages,
which extensively agree (although \texttt{sc2ts} provides a more precise
parent designation).

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. Using Nextclade designations, 15 Pango X lineages are
inferred to have multiple
recombinant origins, such that their samples trace to more than one most recent
recombination node in the ARG. Of these, 11 are cases where the
recombinants are independent rather than nested (i.e., there is no overlap in the list
of descendant samples for each recombination node). Most have a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend and which agrees with the offical Pango designated parent lineages
(see Table~\ref{tab:pango-recombinants}, but note that
in XJ and XU there are too few Pango X samples to decide on a ``main'' clade).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple multiple-origin example, consisting of
the 4 samples labelled XD by Nextclade in the Long ARG. The left hand subgraph
(containing three XD samples, all sampled in France) has an earliest
sample (EPI\_ISL~11222324, dated 2022-02-26), while the right hand
subgraph has a single XD sample (EPI\_ISL~8514045, sampled in Turkey and dated 2021-12-30).
Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
the breakpoints in each case are slightly different: the start of the Omicron insertion
in the French samples has an estimated rightmost position of 21641bp
(and a leftmost of 21619, not shown) with the insertion end occurring at a rightmost position of 25584
(and a leftmost of 25470). By contrast, the Omicron insertion in the Turkish sample
is inferred to have occurred from position 21619-22578 to position 23605-24130.
The breakpoint difference, the different geographical locations, the time between the
samples, and the fact that the two samples differ at 23 nucleotide positions,
suggests that these may indeed represent independent Delta--Omicron recombinants.
The canonical XD definition is based entirely on samples from northern Europe, particularly
France (\url{https://github.com/cov-lineages/pango-designation/issues/444}) so it seems
plausible that the earlier Turkish sample has been mislabelled as XD by Nextclade.
Indeed, GISAID does not label any of these samples XD (see
Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).
Investigation of other multiple-origin examples reveals somewhat similar patterns,
suggesting that most of the simple multiple origin examples are due to incorrect
Pango labelling.

\subsubsection{Nested recombinant origins}
As well as cases where Pango X lineage origins are attributed to independent
recombination events, four Pango X lineages in the Long ARG have Nextclade-designated
samples whose ancestry involves further recombination events
(marked by~\textdagger ~in Table~\ref{tab:pango-recombinants}; the most
complex appears to be XAB). Figure~\ref{fig:complex_origins_graph} plots the earliest example,
XB,  which is present in both the Wide and Long ARGs. The subgraph shows a recombination
between a B.1 sample and B.1.627 sample that leads not only to all the XB-labelled samples
but also to a ``hairball'' of further recombination nodes whose descendants are often not
identified as recombinants by Nextclade (plotted on the left, in blue). A similar
pattern is seen when examining XB in the Wide ARG (see discussion below).
